Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-09-10 pnas/Towards.tex
small editing

@@

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.24\linewidth}
		\centering
		\includegraphics[width=1.1\linewidth]{Einstien_progress_1}
		\caption{$2\times 10^2$ micrographs}
	\end{subfigure} \hfill
	\begin{subfigure}[h]{0.24\linewidth}
		\centering
		\includegraphics[width=1.1\linewidth]{Einstien_progress_4}
		\caption{$2\times 10^3$ micrographs}
	\end{subfigure} \hfill
	\begin{subfigure}[h]{0.24\linewidth}
		\centering
		\includegraphics[width=1.1\linewidth]{Einstien_progress_40}
		\caption{$2\times 10^4$ micrographs}
	\end{subfigure} \hfill
	\begin{subfigure}[h]{0.24\linewidth}
		\centering
		\includegraphics[width=1.1\linewidth]{Einstien_progress_400}
		\caption{$2\times 10^5$ micrographs}
	\end{subfigure}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized randomly. Estimates are obtained from $2\times 10^2,2\times 10^3,2\times 10^4,2\times 10^5$ micrographs (growing across panels), each containing $700$ image occurrences on average.}	
	\vspace{-12pt} 
\end{figure*}

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical SNR. This fact has been recognized early on by the cryo-EM community. In particular, in an influential paper from 1995, Richard Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{myquote}{0.1in}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{myquote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight (below $\sim$50 kDa). 
Joachim Frank voices a similar observation in his 2017 Nobel prize lecture: ``\emph{Using the ribosome as an example, it became clear from the formula we obtained that the single-particle approach to structure research was indeed feasible for molecules of sufficient size: Particle Size > 3/[Contrast$^2$ $\times$ Resolution (as length) $\times$ Critical Electron Dose]}''~\cite{frank2018single}. 
As these two leaders of the cryo-EM community point out, it is impossible to reconstruct small molecules by any of the existing computational pipelines for single particle analysis in cryo--EM, because the particles themselves cannot be picked from the micrographs. Yet, imaging such molecules is important for a number of applications~\cite{scapin2018cryo}

@@

Let $x\in\RL$ be the target signal and let $y\in\RN$ be the observed data, where we assume $N$ is  far larger than $L$. Let  $s \in \{0, 1\}^{N-L+1}$ be a binary signal indicating (with 1's) the starting positions of all occurrences of $x$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
y & =  x \ast s + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. 
While both $x$ and $s$ are unknown, the goal is only to estimate $x$ from $y$.
This is a key difference with other works on  \emph{blind deconvolution}, a longstanding problem arising in a variety of engineering and scientific applications such as astronomy, communication, image deblurring, system identification and optics; see~\cite{jefferies1993restoration,shalvi1990new,ayers1988iterative,abed1997blind}, just to name a few. 
The parameters of the signal $s$ (the locations of its nonzero values) are the \emph{nuisance variables} of the problem. As will be shown next, in a low SNR environment, estimating $s$ is impossible, whereas estimating $x$ is tractable under some conditions on $s$.

>>>>

\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}

@@

\begin{figure*}[h]
	\centering
	\includegraphics[width=1\linewidth]{Einstien_progress_examples}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized randomly. Estimates are obtained from $2\times 10^2,2\times 10^3,2\times 10^4,2\times 10^5$ micrographs (growing across panels from left to right), each containing $700$ image occurrences on average.}	
	\vspace{-12pt} 
\end{figure*}

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical SNR. This fact has been recognized early on by the cryo-EM community. In particular, in an influential paper from 1995, Richard Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{myquote}{0in}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{myquote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight (below $\sim$50 kDa). 
Joachim Frank voices a similar observation in his 2017 Nobel prize lecture: ``\emph{Using the ribosome as an example, it became clear from the formula we obtained that the single-particle approach to structure research was indeed feasible for molecules of sufficient size: Particle Size > 3/[Contrast$^2$ $\times$ Resolution (as length) $\times$ Critical Electron Dose]}''~\cite{frank2018single}. 
As these two leaders of the cryo-EM community point out, it is impossible to reconstruct small molecules by any of the existing computational pipelines for single particle analysis in cryo--EM, because the particles themselves cannot be picked from the micrographs. Yet, imaging such molecules is important for a number of applications~\cite{scapin2018cryo}

@@

Let $x\in\RL$ be the target signal and let $y\in\RN$ be the observed data, where we assume $N$ is  far larger than $L$. Let  $s \in \{0, 1\}^{N-L+1}$ be a binary signal indicating (with 1's) the starting positions of all occurrences of $x$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
y & =  x \ast s + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. 
While both $x$ and $s$ are unknown, the goal is only to estimate $x$ from $y$.
This is a key difference with other works on  \emph{blind deconvolution}, a longstanding problem arising in a variety of engineering and scientific applications such as astronomy, communication, image deblurring, system identification and optics; see~\cite{jefferies1993restoration,shalvi1990new,ayers1988iterative,abed1997blind}
, just to name a few. 
The parameters of the signal $s$ (the locations of its nonzero values) are the \emph{nuisance variables} of the problem. As will be shown next, in a low SNR environment, estimating $s$ is impossible, whereas estimating $x$ is tractable under some conditions on $s$.
